---
title: "Estimation scheme for fitting the stochastic model for the transmission of 2019-nCov in Hubei"
author: "Andrew Tredennick & John Drake"
date: "`r format(Sys.time(), '%B %d, %Y')`"
header-includes:
  - \usepackage{amsmath}
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=FALSE, warning = FALSE, message = FALSE)
```

## Introduction

This document describes a proposed estimation scheme for fitting the stochastic model (described elsewhere) of COVID-19 transmission to data.
We plan to use Approximate Bayesian Computation to estimate unknown parameters.
The posterior distribution of parameters will be explored using MCMC.
While the end goal is estimate parameters given real data, we start by using the model itself to simulate a trajectory of disease transmission and tuning the model fitting process to ensure we can recover known parameters.


## Simulated data

Here we simulate data from the stochastic model using the ```simulators.R``` functions.
The ```simulators.R``` functions are wrappers around the core model functions written by Drake and Rohani.


```{r create-sims}
# First source the necessary functions
source("stochastic-model.R")  # the stochastic disease transmission model
source("simulators.R")  # simulator wrappers
source("style-definitions.R")  # colors

# Time spans for simulation
start <- as.Date("2019-12-01")
today <- Sys.Date()

# Unknown parameters
theta <- list(beta0 = 0.657, beta.factor = 1, sigma = 1/6.4, I0 = 1)

# Known parameters
extraArgs <- list(nstep = NULL, start = start, 
                  today = today, dt = 0.05, w = 40, z = 45, c = 1, 
                  presymptomatic = 0, timesToObs = FALSE, 
                  b = 0.143, a0 = 0.0446, paramNames = names(theta), 
                  nObs = NULL, obsDataDate = NULL)

# Initial conditions
init <- list(S=59002000, E1=0, E2=0, E3=0, E4=0, E5=0, E6=0,
             I1 = NA, I2= 0, I3=0, I4=0, Iu1=0, Iu2=0, Iu3=0, Iu4=0,
             H=0, Ru=0, C=0)
extraArgs$init <- init

# Simulate from the model
res <- mod_wrap(param = log(unlist(theta)), nsim = 2, extraArgs = extraArgs)

matplot(t(res), type = "l", col = col.cases.ci, lty = 1, 
        main = "Simulated trajectories of COVID-19",
        xlab = "Day", ylab = "Number of new cases")
```

We can save one simulated trajectory as "data" for testing of the estimation scheme.

```{r sim-data}
obs_data <- mod_wrap(param = log(unlist(theta)), nsim = 1, extraArgs = extraArgs)
extraArgs$obsData <- obs_data
extraArgs$nObs <- length(obs_data)
plot(obs_data, type='b',  xlab='Day', col = col.cases,
     ylab='Number of new cases', main='Simulated COVID-19 cases in Hubei')
```

## Summary statistics for the ABC

Fitting models using ABC requires calculating a vector of summary statistics ($S$) that summarize the observed time series and simulated trajectories.
The summary statistics ($S_{\text{sim}}$) from simulated time series are compared to the summary statistics of the observed time series ($S_{\text{obs}}$) using a distance function $d$.
To begin with, I used a simple $d$ function that calculates the mean absolute diference across all $N$ summary statistics: $\sum^{N}_{i=1} |S_{i,\text{obs}} - S_{i,\text{sim}}| / N$.

For our particular problem, I used the following summary statistics based on phases of the epidemic that are defined by case numbers.
Specifically, we used a five day moving average of the differences of log observed new cases (plus 10) to define the following phases: day at which the the moving average of log differences exceeds 0.1 ($d_0$) and the day at which the moving average of log differences drops below -0.01 ($d_1$).
The plot below shows the time series of the five-day moving average of the differences of new cases, along with the calculated break points defined above

```{r plot-ma}
ma <- function(x, n = 5){stats::filter(x, rep(1 / n, n), sides = 2)}
stat_times <- ma(diff(log(obs_data+10)), n = 5)
d0 <- min(which(stat_times > 0.1))
d1 <- d0 + min(which(stat_times[d0:length(stat_times)] < -0.01))
par(mfrow = c(1,2))
plot(stat_times, xlab = "Day", ylab = "Moving average of differences", 
     type = "b", cex = 0.5)
abline(h = 0, col = "grey", lty = 2)
abline(v = c(d0, d1), col = "red")
plot(obs_data, type='b',  xlab='Day', col = col.cases,
     ylab='Number of new cases', cex = 0.5)
abline(v = c(d0, d1), col = "red")
```

The summary statistics are as follows:

1. Maximum incidence before day $d_0$ (stuttering chain of tranmission)
2. Maximum incidence between day $d_0$ and day $d_1$ (exponential phase)
3. Maximum incidence after day $d_1$ (decling phase)
4. Cumulative incidence before day $d_0$
5. Cumulative incidence between day $d_0$ and day $d_1$
6. Cumulative incidence after day $d_1$
7. Maximum incidence along the whole trajectory
8. Day of epidemic peak (timing of max incidence)
9. Final size of epidemic (total incidence)

Now we code up calculations for these summary statistics for ```synlik```.

```{r summary-stats}
covid_stats <- function(x, extraArgs, ...) {
  ## obsData is a vector of observed path
  ## x is a M by n.t matrix of paths, each row of 'x' is a replicate
  
  ma <- function(x, n = 5){stats::filter(x, rep(1 / n, n), sides = 2)}
  stat_times <- ma(diff(log(extraArgs$obsData+10)), n = 5)
  d0 <- min(which(stat_times > 0.1))
  d1 <- d0 + min(which(stat_times[d0:length(stat_times)] < -0.01))
  
  obsData <- extraArgs$obsData
  
  stopifnot(is.vector(obsData), length(obsData) != 0)
  if (!is.matrix(x)) x <- matrix(x, 1, length(x))
  
  x <- log(x + 10)  # log transform
  
  # Max incidence
  X0 <- (apply(x, 1, function(x) max(x[1:(d0-1)])))
  X0 <- cbind(X0, (apply(x, 1, function(x) max(x[d0:d1]))))
  X0 <- cbind(X0, (apply(x, 1, function(x) max(x[(d1+1):length(x)]))))
  
  # Cumulative incidence
  X0 <- cbind(X0, (apply(x, 1, function(x) sum(x[1:(d0-1)]))))
  X0 <- cbind(X0, (apply(x, 1, function(x) sum(x[d0:d1]))))
  X0 <- cbind(X0, (apply(x, 1, function(x) sum(x[(d1+1):length(x)]))))
  
  # Max along whole trajectory
  X0 <- cbind(X0, apply(x, 1, max))
  
  # Day of max
  X0 <- cbind(X0, apply(x, 1, function(x) which.max(x)))
  
  # Exponential increase and decline
  X0 <- cbind(X0, apply(x, 1, max) / (apply(x, 1, function(x) which.max(x)) - d0))
  
  # Linear regression coefs
  X0 <- cbind(X0, apply(x, 1, function(x) as.numeric(coef(lm(x[1:(d0-1)] ~ seq_along(x[1:(d0-1)])))[2])))
  X0 <- cbind(X0, apply(x, 1, function(x) as.numeric(coef(lm(x[d0:d1] ~ seq_along(x[d0:d1])))[2])))
  X0 <- cbind(X0, apply(x, 1, function(x) as.numeric(coef(lm(x[(d1+1):length(x)] ~ seq_along(x[(d1+1):length(x)])))[2])))
  
  # Final epidemic size
  X0 <- cbind(X0, apply(x, 1, sum))
  return(X0)
}
```


Here are example statistics from a few simulations.

```{r ex-stats}
summaries <- covid_stats(x = res, extraArgs = extraArgs)
print(summaries)
```

And the statistics from the "data".

```{r obs-stats}
obs_summaries <- covid_stats(x = obs_data, extraArgs = extraArgs)
print(obs_summaries)
```

Last, we define a distance function $d$.
It is important to know how $d$ might vary due to model stochasticity alone because we have to set a threshold value ($\varepsilon$) that $d$ must fall below for acceptance of a parameter vector in the ABC-MCMC algorithm.
So, I simulated 200 trajectories from the same parameter values and calculated the distances of the summary statistics.
The resulting histogram indicates an initial $\varepsilon \approx 10$.

```{r dist}
d <- function(xsim, xobs) {
  mean(abs(xobs - xsim))
}

d(summaries[2,], obs_summaries)

dsims <- 2
dist <- numeric(dsims)
for(i in 1:dsims) {
  res <- mod_wrap(param = log(unlist(theta)), nsim = 1, extraArgs = extraArgs)
  summaries <- covid_stats(x = res, extraArgs = extraArgs)
  dist[i] <- d(summaries, obs_summaries)
}
hist(dist, main = "Distribution of statistic distances at known parameters",
     xlab = "d")
abline(v = median(dist), col = "red")

```

## ABC Algorithm
```{r abc}
get_prior_d <- function(theta, theta_prime) {
  d_theta <- dnorm(theta[1], log(0.657), 0.5) +
             dnorm(theta[2], log(1), 0.5) +
             dnorm(theta[3], log(1/6,4), 0.5) +
             dunif(theta[4], -5, 5)
  
  d_theta_prime <- dnorm(theta_prime[1], log(0.657), 0.5) +
                   dnorm(theta_prime[2], log(1), 0.5) +
                   dnorm(theta_prime[3], log(1/6,4), 0.5) +
                   dunif(theta_prime[4], -5, 5)
  return(min(1, exp(d_theta_prime - d_theta)))
}


run_abc <- function(nburn, nsamp, epsilon, f_stats, f_distance, f_model, 
                    init_theta, obs_data, ...) {
  results <- matrix(nrow = nsamp, ncol = length(init_theta))
  i <- 0
  theta <- init_theta
  while(i < (nburn+nsamp)) {
    theta_prime <- rnorm(length(theta), theta, sd = c(0.08, 0.08, 0.08, 0.08))
    p <- f_prior(as.numeric(theta), theta_prime)
    if(rbinom(1, 1, p) == 1) {
      m <- f_model(param = theta_prime, nsim = 1, extraArgs = extraArgs)
      mS <- f_stats(x = m, extraArgs = extraArgs)
      mO <- f_stats(x = obs_data, extraArgs = extraArgs)
      dtest <- f_distance(mS, mO)
      if(dtest < epsilon) {
        theta <- theta_prime
      }
    }
#     propsal = c
#     accept with p exp(prior(propsal) – prior(current))
#     if accept do ABC step
    i <- i + 1
    if(i > nburn) results[(i - nburn), ] <- theta
  }
  return(results)
}
```

```{r test}
init_theta <- log(c(beta0 = 0.5, beta.factor = 2, sigma = 0.2, I0 = 5))
out_mcmc <- run_abc(nburn = 100, nsamp = 500, epsilon = 10, f_stats = covid_stats, 
                    f_distance = d, f_model = mod_wrap, f_prior = get_prior_d,
                    init_theta = init_theta, obs_data = obs_data)

# Plot the chains
par(mfrow = c(2, 2))
plot(out_mcmc[,1], type = "l", xlab = "Iteration",
     ylab = expression(log(beta)), las = 1)
abline(h = log(as.numeric(theta[1])), col = "red")
plot(out_mcmc[,2], type = "l", xlab = "Iteration",
     ylab = expression(log(xi)), las = 1)
abline(h = log(as.numeric(theta[2])), col = "red")
plot(out_mcmc[,3], type = "l", xlab = "Iteration",
     ylab = expression(log(sigma)), las = 1)
abline(h = log(as.numeric(theta[3])), col = "red")
plot(out_mcmc[,4], type = "l", xlab = "Iteration",
     ylab = expression(log(I[0])), las = 1)
abline(h = log(as.numeric(theta[4])), col = "red")
```



